#This R markdown provide a guided construction of Species Distribution Model of 
#marine organisms using the SSDM package. 

#Step 1: Loading the library we need to use 
```{r}
library(SSDM)
```


#Step 2: Specifiy the path to the present-day environmental data 
#--------------------------------------------------------------------------------------
#Note: We are using a Jupyter Hub with all the files pre-loaded in the same folder
#If you are going to use this file offline on your own device, please make sure you 
#enter the appropriate path. 
#To find a path on PC, right click on the file and select properties, the path is the Location 
#For PC users, make sure you change the backward slash (\) to forward slash (/) 
# To find a path on Mac, Open Finder.Click View in the top menu bar.Select Show Path Bar. 
#--------------------------------------------------------------------------------------

```{r}
Env_present_path <- "C:/Data/UW_work/teaching/ocn431/Fall2026/SDM_Env_data/Present_day"
```

#Step 3: Load the present-day envriomental data you have selected
#----------------------------------------------------------------------------------------
#Make sure each file name is entered exactly as is and that each file name has 
#paranethesis around them and comma separating each file. 
#--------------------------------------------------------------------------------------
```{r}
Env0 <- load_var(path = Env_present_path, files = c("2000_2019_temp_mean.nc","2000_2019_pH_min.nc"))
```

#Plot to check the environmental data looks OK 
```{r}
plot(Env0)
```



#Step 4:  Specifiy the path to the present-day occurence data
#--------------------------------------------------------------------------------------
#Note: We are using a Jupyter Hub with all the files pre-loaded in the same folder
#If you are going to use this file offline on your own device, please make sure you 
#enter the appropriate path. 
#--------------------------------------------------------------------------------------

```{r}
occ_path <- "C:/Data/UW_work/teaching/ocn431/Fall2026/SDM_Occ_data"
```


#Step 5: Load the Occurence Data
#--------------------------------------------------------------------------------------
#Note: For this exercise, the occurence data from GBiF has been vetted and trimmed for you
#You just have to specificy the correct file. Do not change the headers of columns,
# unless you adjust the naming accordingly in the command below
#--------------------------------------------------------------------------------------
```{r}
Occ0 <- load_occ(
path = occ_path, Env, file = "spurp2.csv",
Xcol = "decimalLongitude", Ycol = "decimalLatitude", Spcol = "species", sep = ",")
```


#Step 6: Run the SSDM model 
```{r}
SDM_now <- modelling('GLM', Occ0, Env0, Xcol = "decimalLongitude", Ycol = "decimalLatitude", verbose = FALSE)
```

#Step 7: Plot the modelled habitat suitability
```{r}
plot(SDM_now@projection, main = 'Present-day habitat suitability with GLM algorithm')
```
#Note: This is a good point to re-evaluate if you need to change the environemtnal variables you have chosen
#If the suitability map does not make sense, consider adjusting. 

#Step 8: Check model quality and relative importance of the environmental variables 
```{r}
SDM_now@evaluation
SDM_now@variable.importance
```

#Step 9: Save the present-day suitability model so that you can visualize and compute changes later 
```{r}
proj_now <- SDM_now@projection
```

#Now that you are ready to look forward in time and see how suitability change with anthropogenic climate change 
#Step 10: Specify the path to the future environmental data 
```{r}
Env_future_path <- "C:/Data/UW_work/teaching/ocn431/Fall2026/SDM_Env_data/Future"
```

#Step 11: Load the future environment data 
```{r}
Env_new <- load_var(path = future_path, files = c("future_temp_mean.nc","future_pH_min.nc"))
```

#Don't forget to plot the future environmental data to check what are the changes 
```{r}
plot(Env_new)
```

#Step 12: Project future habtiat suitability based on the relationships developed based on present-day data 
```{r}
 SDM_forecast <- project(SDM, Env_new)	
```

#Plot and save the future projection
```{r}
proj_future <- SDM_forecast@projection
plot(SDM_forecast@projection, main = 'SDM forecast with GLM algorithm
under changing climate')

```



#Step 13: Evaulate the change in suitability 

```{r}
#Some trick to remove the land area in the map
ocean_mask <- reclassify(proj_now, cbind(-Inf, Inf, 1))
proj_area <- area(proj_now) * ocean_mask

#To determine the change in habitable area over time, you need to first decide on a threshold. This threshold correspond to the probability of presence according to your model. For instance, a threshold of 0.1 would mean the model suggests there is a 10% chance for the species of interest to be present at the specific cell (location). Then you can apply the threshold to compute the habitable area for now and future.
thres <- 0.1

#Compute the present-day and future habtiatble area 
livable_now <- cellStats(proj_area * (proj_now > thres), sum)
livable_future <- cellStats(proj_area * (proj_future > thres), sum)

#You can visualize the change in suitability by plotting the difference
proj_diff <- proj_future - proj_now 
plot(proj_diff, main = "Change in suitability probability")

```

